---
title: "A Practical scFPCDE Workflow"
subtitle: "From expression filtering to FPCA tuning and permutation inference"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{A Practical scFPCDE Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  fig.align = "center",
  dpi = 96
)
options(width = 90)
```

```{r setup}
library(scFPCDE)
set.seed(2026)
```

# What scFPCDE does

`scFPCDE` identifies genes whose expression functions vary along single-cell
pseudotime. It treats the collection of gene trajectories as functional data,
estimates a shared functional principal component analysis (FPCA) basis, and
uses gene-level FPCA scores to construct permutation tests.

The package is most useful after a trajectory or lineage has already been
defined. It does **not** infer pseudotime, select a lineage, normalize counts, or
correct batch effects. Those steps should be completed before the expression
matrix and pseudotime vector are supplied to `scFPCDE`.

The main functions form one workflow:

```{r function-map, echo=FALSE}
function_map <- data.frame(
  Step = c(
    "1. Filter", "2. Tune", "3. Fit", "4. Test", "5. Explore", "6. Visualize"
  ),
  Function = c(
    "scFPCDE_filter_genes()",
    "scFPCDE_tune_fpca()",
    "scFPCDE_fit_fpca() or scFPCDE_run()",
    "scFPCDE_D_test() and scFPCDE_F_test()",
    "scFPCDE_fpc_scores()",
    "scFPCDE_gene_curves()"
  ),
  Purpose = c(
    "Remove genes with low detection frequency",
    "Choose spline-basis size and roughness penalty by GCV",
    "Estimate smooth trajectories and shared FPCs",
    "Evaluate gene-level evidence against a permutation null",
    "Extract, select, or transform gene-level FPC scores",
    "Compare observed expression with fitted trajectories"
  ),
  check.names = FALSE
)
knitr::kable(function_map, align = c("l", "l", "l"))
```

`scFPCDE_run()` is the recommended high-level entry point. It orders cells by
pseudotime, preprocesses expression, fits an initial FPCA model, refines the
eigenfunctions using genes with the strongest scores, and runs the requested
permutation tests.

# Input contract

The required inputs are:

- `yt`: a finite numeric matrix with **cells in rows** and **genes in columns**;
- `tt`: one finite pseudotime value per row of `yt`; and
- gene identifiers in `colnames(yt)`.

Pseudotime can use any finite scale. The methods depend on its ordering and
relative spacing, not on a specific range. Missing values should be resolved
before fitting. The same biological lineage should not contain unrelated cells
whose pseudotime values are not comparable.

Important modeling choices are:

- `center = TRUE` removes each gene's mean and focuses the analysis on shape;
- `scale = TRUE` additionally gives genes unit variance, which can make weak and
  strong genes contribute more equally;
- `L` is the number of functional principal components retained;
- `nbasis` controls the maximum flexibility of the B-spline representation;
- `r_pen` penalizes curvature, with larger values producing smoother fits;
- `topvarper` is the proportion of high-score genes used to refine the shared
  eigenfunctions; and
- `fpc_varmax = TRUE` rotates the retained FPCs for interpretability without
  changing their retained subspace.

# A known-truth example

The bundled `scFPCDE_simdata` object contains 1,000 cells and 500 genes. Genes
1--100 were generated with pseudotime-dependent trajectories, while genes
101--500 are null. This known status lets us demonstrate a compact
single-dataset simulation evaluation. It is useful for learning the workflow,
but it is not a claim about performance across all data-generating mechanisms.

```{r data}
data(scFPCDE_simdata)

# Use the complete bundled simulation: no cell or gene preselection.
yt <- scFPCDE_simdata$yt
tt <- scFPCDE_simdata$tt
cell_cluster <- scFPCDE_simdata$clusters

known_signal <- setNames(
  seq_len(ncol(yt)) <= 100,
  colnames(yt)
)

data.frame(
  Cells = nrow(yt),
  Genes = ncol(yt),
  Known_signals = sum(known_signal),
  Known_nulls = sum(!known_signal),
  Pseudotime_min = min(tt),
  Pseudotime_max = max(tt)
)
```

The complete bundled dataset is already ordered by pseudotime.
`scFPCDE_run()` also orders its inputs internally, so an unsorted input is
acceptable.

# Filtering low-detection genes

`scFPCDE_filter_genes()` counts the nonzero observations for each gene and
retains genes above a user-specified quantile of those counts. It is intended
for sparse count-like or log-expression matrices with heterogeneous detection.

The bundled simulation is continuous and contains no zeros. Because all 500
genes therefore have the same detection count, the strict quantile rule cannot
distinguish them and should not be applied to the original matrix. To
demonstrate the requested 5% filter without pretending otherwise, the next
chunk creates a documented didactic copy in which 50 known-null genes are
observed in only 10 cells. The original, unmodified full simulation is used for
the FPCA analysis that follows.

```{r filtering}
yt_sparse_demo <- yt
sparse_gene_columns <- 451:500

yt_sparse_demo[, sparse_gene_columns] <- 0
yt_sparse_demo[1:10, sparse_gene_columns] <-
  yt[1:10, sparse_gene_columns]

detection_count <- colSums(yt_sparse_demo != 0)
filter_quantile <- 0.05
detection_threshold <- unname(quantile(detection_count, filter_quantile))

genes_kept <- scFPCDE_filter_genes(
  yt_sparse_demo,
  qz = filter_quantile
)

data.frame(
  Input_genes = ncol(yt_sparse_demo),
  Detection_threshold = detection_threshold,
  Retained_genes = length(genes_kept),
  Removed_genes = ncol(yt_sparse_demo) - length(genes_kept)
)
```

```{r filtering-plot, fig.cap="Genes above the 5th-percentile detection threshold are retained. A logarithmic y-axis keeps both low- and high-detection genes visible."}
sorted_detection <- sort(unname(detection_count))
gene_rank <- seq_along(sorted_detection)
gene_status <- sorted_detection > detection_threshold

plot(
  gene_rank,
  sorted_detection,
  type = "n",
  log = "y",
  ylim = c(max(1, min(sorted_detection) / 2), max(sorted_detection) * 1.2),
  xlab = "Genes ordered by detection count",
  ylab = "Cells with nonzero expression (log scale)",
  main = "Quantile-based detection filter"
)
abline(h = detection_threshold, lty = 2, lwd = 2, col = "#D95F02")
points(
  gene_rank,
  sorted_detection,
  pch = 16,
  cex = 0.65,
  col = ifelse(gene_status, "#2C7FB8", "grey65")
)
legend(
  "left",
  legend = c("Retained", "Removed", "Threshold"),
  pch = c(16, 16, NA),
  lty = c(NA, NA, 2),
  col = c("#2C7FB8", "grey65", "#D95F02"),
  bty = "o",
  bg = "white",
  box.col = "grey85",
  cex = 0.9
)
```

In a real workflow, the filtered matrix would be constructed with:

```{r filtered-matrix}
yt_filtered <- yt_sparse_demo[, genes_kept, drop = FALSE]
dim(yt_filtered)
```

If every gene is detected in exactly the same number of cells, a strict
quantile filter cannot distinguish genes and should be skipped. Filtering also
changes the multiple-testing universe, so its rule should be chosen before
examining gene-level p-values.

# Tuning the roughness penalty with a fixed basis

Two parameters work together to determine trajectory smoothness:

- `nbasis` is the number of B-spline basis functions. A larger value permits
  more local variation but does not force a wiggly result.
- `r_pen` is the penalty on the integrated squared second derivative. A larger
  penalty suppresses curvature and favors smoother trajectories.

For this example, the B-spline basis is fixed at 30 functions and
`scFPCDE_tune_fpca()` evaluates only the roughness penalty by mean generalized
cross-validation (GCV). Fixing the basis makes the diagnostic a single curve:
lower GCV indicates a better balance between fit and smoothness for the genes
used during tuning.

For reproducibility, this vignette uses sequential evaluation. The penalty grid
spans six orders of magnitude and contains the observed minimum.

```{r tune}
fixed_nbasis <- 30L
roughness_grid <- 10^seq(-7, -1, by = 0.5)

tuning <- scFPCDE_tune_fpca(
  yt = yt,
  tt = tt,
  L = 2,
  nbasis_range = fixed_nbasis,
  r_pen_range = roughness_grid,
  topvarper = 0.25,
  ncores = 1
)

tuning_table <- transform(
  tuning$parameter_grid,
  Mean_GCV = tuning$GCV_results,
  Roughness_penalty = format(r_pen, scientific = TRUE, digits = 2)
)

knitr::kable(
  tuning_table[, c("nbasis", "Roughness_penalty", "Mean_GCV")],
  digits = c(0, 6, 6),
  col.names = c("B-spline basis functions", "Roughness penalty", "Mean GCV")
)
```

```{r tuning-plot, fig.cap="Mean generalized cross-validation over the roughness-penalty grid with the B-spline basis fixed at 30 functions. The orange point and dashed line identify the selected penalty."}
ggplot2::ggplot(
  tuning_table,
  ggplot2::aes(x = r_pen, y = Mean_GCV)
) +
  ggplot2::geom_line(
    color = "#0072B2",
    linewidth = 1
  ) +
  ggplot2::geom_point(
    color = "#0072B2",
    size = 2.2
  ) +
  ggplot2::geom_vline(
    xintercept = tuning$best_r_pen,
    linetype = 2,
    color = "#D55E00",
    linewidth = 0.8
  ) +
  ggplot2::geom_point(
    data = tuning_table[tuning_table$r_pen == tuning$best_r_pen, , drop = FALSE],
    color = "#D55E00",
    size = 3
  ) +
  ggplot2::scale_x_log10(
    breaks = 10^seq(-7, -1),
    labels = function(x) format(x, scientific = TRUE)
  ) +
  ggplot2::labs(
    x = "Roughness penalty (log10 scale)",
    y = "Mean GCV",
    title = "Roughness-penalty tuning with 30 basis functions"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

```{r selected-tuning}
data.frame(
  L = tuning$best_L,
  Fixed_nbasis = fixed_nbasis,
  r_pen = tuning$best_r_pen,
  Mean_GCV = tuning$GCV
)
```

Here, the fixed 30-function basis gives a clear interior GCV minimum over the
candidate penalties. For a final analysis, a minimum on either boundary is a
reason to expand the penalty grid and check whether the choice is stable. The
fixed basis and penalty grid should also be revisited when the lineage,
normalization, or cell density changes.

`L` is intentionally specified rather than selected by GCV. Two or three FPCs
are common starting points. Increase `L` only when additional components explain
meaningful structured variation and yield stable results; excessive harmonics
can model noise.

# Run the complete pipeline with 100 permutations

The next call uses the tuned parameters and runs both available test
statistics. One hundred permutations provide a meaningful, quick tutorial.
Final inferential analyses should normally use at least 1,000 permutations and
should examine stability across random seeds.

`ncores = 1` makes the seeded permutation sequence reproducible across the
platforms used to build this vignette. Parallel execution can be enabled for a
large analysis, but its random-number behavior should be recorded and checked.

```{r run}
set.seed(2026)

result <- scFPCDE_run(
  yt = yt,
  tt = tt,
  L = tuning$best_L,
  nbasis = fixed_nbasis,
  r_pen = tuning$best_r_pen,
  topvarper = 0.25,
  center = TRUE,
  scale = FALSE,
  n_perm = 100,
  ncores = 1,
  use_FPC_F = TRUE,
  fpc_varmax = TRUE
)

names(result)
```

The two gene-level tests answer related questions:

- The **D-statistic** is the squared norm of a gene's retained FPCA scores. It
  measures how far the fitted trajectory lies from the null origin in the
  retained score space.
- The **F-statistic** compares residual variation under the null and FPCA
  models. It asks whether the fitted functional representation improves on the
  null model.

Both result tables report raw permutation `p_value` and Benjamini--Hochberg
`q_value`. The q-values should be used when controlling false discoveries over
the tested genes.

# Inspect the fitted FPCA representation

The refined `fpca_result` stores gene scores, evaluated eigenfunctions, fitted
expression, variance proportions, and underlying `fda` objects.

```{r fpca-components}
fpca_summary <- data.frame(
  Component = paste0("FPC", seq_along(result$fpca_result$PEV)),
  Proportion = result$fpca_result$PEV,
  Cumulative = cumsum(result$fpca_result$PEV)
)
knitr::kable(fpca_summary, digits = 3)
```

```{r eigenfunctions, fig.cap="The two retained eigenfunctions describe the dominant trajectory shapes shared across genes."}
matplot(
  tt,
  result$fpca_result$eigenfunctions,
  type = "l",
  lty = 1,
  lwd = 2.2,
  col = c("#0072B2", "#D55E00"),
  xlab = "Pseudotime",
  ylab = "Eigenfunction value",
  main = "Retained functional principal components"
)
abline(h = 0, lty = 3, col = "grey65")
legend(
  "topright",
  legend = paste0("FPC", seq_len(ncol(result$fpca_result$eigenfunctions))),
  col = c("#0072B2", "#D55E00"),
  lty = 1,
  lwd = 2.2,
  bty = "n"
)
```

## Extracting and transforming FPC scores

`scFPCDE_fpc_scores()` provides one consistent way to retrieve scores from
either `scFPCDE_fit_fpca()` or `scFPCDE_run()`. Raw scores must be used to
calculate the D-statistic, permutation null, distances, or score-space
boundaries. Two optional transformations are useful for visualization:

- `transform = "signed_log10"` applies
  $\operatorname{sign}(x)\log_{10}(1 + |x|/s)$. It preserves zero, sign, and
  score-space quadrant while compressing extreme values. `log_scale = 1`
  exactly reproduces the manuscript plotting transformation.
- `transform = "standardize"` centers every selected FPC and divides by its
  sample standard deviation. This puts components on comparable visual scales,
  but distances are then expressed in standard-deviation units rather than the
  fitted FPCA units.

```{r transform-scores}
scores_raw <- scFPCDE_fpc_scores(result, components = 1:2)
scores_log <- scFPCDE_fpc_scores(
  result,
  components = 1:2,
  transform = "signed_log10"
)
scores_z <- scFPCDE_fpc_scores(
  result,
  components = c("FPC1", "FPC2"),
  transform = "standardize"
)

data.frame(
  Scale = c("Raw", "Signed log10", "Standardized"),
  FPC1_min = c(min(scores_raw[, 1]), min(scores_log[, 1]), min(scores_z[, 1])),
  FPC1_max = c(max(scores_raw[, 1]), max(scores_log[, 1]), max(scores_z[, 1]))
)
```

The three panels below contain the same genes. Compression makes the central
score cloud easier to see when a few genes have extreme coordinates, whereas
standardization makes variation along FPC1 and FPC2 visually comparable. In
this known-truth example, orange marks signal genes and grey marks null genes.
In real data, colors could instead represent a significance category or
external gene annotation.

```{r score-space, fig.width=7.5, fig.height=3.2, fig.cap="Raw, signed-log10-compressed, and standardized views of the same gene-level FPC scores. Transformations change only the display coordinates."}
score_truth <- known_signal[rownames(scores_raw)]
score_colors <- ifelse(score_truth, "#D55E00", "grey72")

old_par <- par(mfrow = c(1, 3), mar = c(4.1, 4.1, 2.6, 0.8))
score_views <- list(
  Raw = scores_raw,
  `Signed log10` = scores_log,
  Standardized = scores_z
)
for (view_name in names(score_views)) {
  score_view <- score_views[[view_name]]
  plot(
    score_view[, 1], score_view[, 2],
    pch = 16,
    cex = 0.72,
    col = score_colors,
    xlab = "FPC1",
    ylab = "FPC2",
    main = view_name
  )
  abline(h = 0, v = 0, lty = 3, col = "grey60")
}
par(old_par)
legend(
  "top",
  legend = c("Known signal", "Known null"),
  col = c("#D55E00", "grey72"),
  pch = 16,
  horiz = TRUE,
  bty = "n"
)
```

# Mini simulation evaluation

Attach the known status only for this simulation and rank genes by the D-test.
This is an oracle evaluation: a real dataset never supplies `Known_signal`.

```{r result-table}
d_result <- transform(
  result$D_test_result,
  Known_signal = unname(known_signal[ID])
)
d_result <- d_result[order(
  d_result$q_value,
  d_result$p_value,
  -d_result$D_obs
), ]

knitr::kable(
  head(d_result, 12),
  digits = c(NA, 3, 4, 4, NA),
  row.names = FALSE,
  caption = "Top genes ranked by the scFPCDE D-test"
)
```

```{r method-summary}
summarize_test <- function(test_result, statistic_name) {
  truth <- unname(known_signal[test_result$ID])
  discovered <- test_result$q_value < 0.05
  true_positive <- sum(discovered & truth)
  false_positive <- sum(discovered & !truth)

  data.frame(
    Test = statistic_name,
    Discoveries = sum(discovered),
    True_positives = true_positive,
    False_positives = false_positive,
    Sensitivity = true_positive / sum(truth),
    Empirical_FDR = false_positive / max(1, sum(discovered))
  )
}

simulation_summary <- rbind(
  summarize_test(result$D_test_result, "D-statistic"),
  summarize_test(result$F_test_result, "F-statistic")
)

knitr::kable(
  simulation_summary,
  digits = c(NA, 0, 0, 0, 3, 3),
  row.names = FALSE,
  caption = "Known-truth summary at q-value < 0.05"
)
```

```{r pvalue-distributions, fig.width=7, fig.height=3.6, fig.cap="Permutation p-value distributions for the D- and F-statistics."}
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
hist(
  result$D_test_result$p_value,
  breaks = seq(0, 1, length.out = 21),
  col = "#9ECAE1",
  border = "white",ylim = c(0,120),
  main = "D-test",
  xlab = "Permutation p-value"
)
hist(
  result$F_test_result$p_value,
  breaks = seq(0, 1, length.out = 21),
  col = "#FDD0A2",
  border = "white",
  main = "F-test", ylim = c(0,120),
  xlab = "Permutation p-value"
)
par(old_par)
```

The summary above shows complete sensitivity for both tests in this full-data
run. The D-test selects no known nulls, whereas the F-test selects two. These
numbers should not be generalized beyond this demonstration. A formal
simulation study would repeat many independently generated datasets across
sample sizes, signal shapes, dropout levels, and pseudotime uncertainty.

# Plot observed and fitted gene trajectories

`scFPCDE_run()` centers genes internally by default, so the observations passed
to `scFPCDE_gene_curves()` should receive the same transformation. The bundled
simulation rows are already in increasing pseudotime order, so the fitted and
observed rows are aligned.

The plot below combines four high-ranking known signals with two of the
highest-ranking known nulls. This makes both clear functional structure and
near-null behavior visible.

```{r gene-curves, fig.width=8, fig.height=5.8, fig.cap="Observed centered expression and FPCA-fitted curves for four known signals and two known null genes."}
yt_centered <- scale(yt, center = TRUE, scale = FALSE)

signal_genes <- head(d_result$ID[d_result$Known_signal], 4)
null_genes <- head(d_result$ID[!d_result$Known_signal], 2)
genes_to_plot <- c(signal_genes, null_genes)

cell_stage <- factor(
  cell_cluster,
  levels = c("0", "1", "2"),
  labels = c("Early", "Middle", "Late")
)

cluster_palette <- c(
  Early = "#0072B2",
  Middle = "#009E73",
  Late = "#D55E00"
)

scFPCDE_gene_curves(
  tt = tt,
  yt = yt_centered,
  yt_fit = result$fpca_result$xt_hat,
  cell_cluster = cell_stage,
  subset = genes_to_plot,
  nrow = 2,
  ncol = 3,
  point_size = 0.55,
  point_alpha = 0.55,
  curve_linewidth = 1,
  cluster_colors = cluster_palette,
  title = "Known signals and nulls along simulated pseudotime"
)
```

For a small number of genes, trajectories can instead be overlaid:

```{r overlay-example, eval=FALSE}
scFPCDE_gene_curves(
  tt = tt,
  yt = yt_centered,
  yt_fit = result$fpca_result$xt_hat,
  cell_cluster = cell_stage,
  subset = signal_genes[1:3],
  facet_genes = FALSE,
  cluster_colors = cluster_palette
)
```

# Using the lower-level functions

The wrapper is convenient, but each analysis component is exported. This is
useful when the same FPCA fit should support custom diagnostics or when D- and
F-tests need different permutation counts.

```{r lower-level, eval=FALSE}
yt_centered <- scale(yt, center = TRUE, scale = FALSE)

fit <- scFPCDE_fit_fpca(
  yt = yt_centered,
  tt = tt,
  L = tuning$best_L,
  nbasis = fixed_nbasis,
  r_pen = tuning$best_r_pen,
  fpc_varmax = TRUE
)

set.seed(2026)
d_test <- scFPCDE_D_test(
  yt = yt_centered,
  tt = tt,
  fpca_result = fit,
  n_perm = 1000,
  ncores = 1
)

set.seed(2026)
f_test <- scFPCDE_F_test(
  yt = yt_centered,
  tt = tt,
  fpca_result = fit,
  n_perm = 1000,
  ncores = 1
)
```

Unlike `scFPCDE_run()`, the lower-level fit does not perform the two-stage
high-score refinement. Use the wrapper when that complete workflow is desired.

# Template for a real analysis

The following non-evaluated template summarizes the recommended progression.

```{r real-template, eval=FALSE}
# expression: cells x genes, normalized and finite
# pseudotime: one value per cell
# cell_state: one cluster or state label per cell

genes_kept <- scFPCDE_filter_genes(expression, qz = 0.05)
expression_use <- expression[, genes_kept, drop = FALSE]

fixed_nbasis <- 30L

tuning <- scFPCDE_tune_fpca(
  expression_use,
  pseudotime,
  L = 2,
  nbasis_range = fixed_nbasis,
  r_pen_range = 10^seq(-7, -1, by = 0.5),
  topvarper = 0.10,
  ncores = 1
)

set.seed(2026)
analysis <- scFPCDE_run(
  expression_use,
  pseudotime,
  L = tuning$best_L,
  nbasis = fixed_nbasis,
  r_pen = tuning$best_r_pen,
  topvarper = 0.10,
  n_perm = 1000,
  ncores = 1,
  use_FPC_F = TRUE
)

# Use raw scores for inference and transformed scores only for display.
scores_raw <- scFPCDE_fpc_scores(analysis, transform = "none")
scores_for_plot <- scFPCDE_fpc_scores(
  analysis,
  components = 1:2,
  transform = "signed_log10"
)

significant_genes <- subset(
  analysis$D_test_result,
  q_value < 0.05
)
```

Before interpreting a final analysis:

1. inspect detection rates and the number of genes retained;
2. verify that the tuning optimum is not an unexplored grid boundary;
3. inspect eigenfunctions, variance proportions, scores, and fitted curves;
4. repeat permutation inference with another seed to assess stability;
5. consider pseudotime uncertainty and repeat the analysis across plausible
   trajectories when possible;
6. report the tested lineage, preprocessing, filtering rule, tuning grid,
   `L`, `topvarper`, permutation count, seed, and package version; and
7. save the complete result and tuning objects rather than only a list of
   significant genes.

```{r save-analysis, eval=FALSE}
saveRDS(
  list(
    tuning = tuning,
    result = analysis,
    seed = 2026,
    package_version = as.character(packageVersion("scFPCDE"))
  ),
  file = "scFPCDE_analysis.rds"
)
```

# Reproducibility information

```{r session-info}
sessionInfo()
```
