---
title: "scFPCDE Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scFPCDE Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(scFPCDE)
library(dplyr)
library(tidyr)
```

## Overview

The `scFPCDE` package provides a statistical framework to detect differential gene expression along pseudotime in single-cell data using Functional Principal Component Analysis (FPCA).

This vignette walks through a complete pipeline using a simulated dataset.

## Load Simulated Dataset

We load a simulated dataset with 1000 cells × 500 genes:
- 100 DE genes are generated as linear combinations of 3 smooth basis functions.
- 400 genes are random noise.
- Cells are assigned to three clusters based on pseudotime.

```{r simulate}
data("scFPCDE_simdata")
sim <- scFPCDE_simdata
yt <- sim$yt
tt <- sim$tt
clusters <- sim$clusters
```

## Run Differential Expression Analysis

We apply the main `scFPCDE_run()` wrapper, which:
- Smooths gene expression curves
- Performs FPCA
- Computes both D-statistic and F-statistic via permutation tests

```{r run}
res <- scFPCDE_run(yt, tt)
```

## Visualize Smoothed Gene Expression Curves

We use `scFPCDE_gene_curves()` to visualize the fitted FPCA curves:
- `yt_fit` is the FPCA-reconstructed matrix
- `cell_cluster` is used to color points by cluster
- We subset the top 12 genes for clarity

```{r plot}
scFPCDE_gene_curves(
  tt = tt,
  yt = yt,
  yt_fit = res$fpca_result$xt_hat,
  cell_cluster = clusters,
  subset = 1:12,
  facet_genes = TRUE,
  nrow = 3,
  ncol = 4
)
```

## View Statistical Test Results

```{r results}
head(res$D_test_result)
head(res$F_test_result)
```

## Conclusion

This example demonstrates how `scFPCDE` can detect genes with smooth, cluster-driven variation over pseudotime. The simulation is customizable, and the full pipeline is modular for real-world data.
