---
title: "Getting Started with scFPCDE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with scFPCDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(scFPCDE)
```

## Overview

`scFPCDE` tests whether gene-expression trajectories vary along single-cell
pseudotime. The main workflow smooths centered expression curves with
functional principal component analysis (FPCA), refines the eigenfunctions
using genes with the strongest FPCA scores, and evaluates gene-level statistics
against a pooled permutation distribution.

This vignette uses a small subset of the bundled simulation and only 25
permutations so that it builds quickly. Analyses intended for inference should
use substantially more permutations, such as the default `n_perm = 1000`.

## Example data

```{r simulate}
data(scFPCDE_simdata)

# Retain 20 DE genes and 40 null genes across 200 cells.
gene_index <- c(1:20, 101:140)
cell_index <- seq(1, length(scFPCDE_simdata$tt), length.out = 200)
cell_index <- unique(round(cell_index))

yt <- scFPCDE_simdata$yt[cell_index, gene_index]
tt <- scFPCDE_simdata$tt[cell_index]
clusters <- scFPCDE_simdata$clusters[cell_index]

dim(yt)
range(tt)
```

Rows of `yt` are cells and columns are genes. `tt` must contain one finite
pseudotime value per cell, and gene names must be present and unique. The wrapper
orders cells by pseudotime and centers genes by default, so the input should not
be centered a second time before calling the wrapper.

## Optional parameter tuning

`scFPCDE_tune_fpca()` evaluates combinations of B-spline basis sizes and
smoothing penalties by generalized cross-validation. A representative tuning
call is shown below. It is not evaluated while building this vignette because a
larger grid can be computationally expensive.

```{r tune, eval=FALSE}
tuning <- scFPCDE_tune_fpca(
  yt = yt,
  tt = tt,
  L = 2,
  nbasis_range = c(20, 30, 40),
  r_pen_range = c(1e-4, 1e-3, 1e-2),
  topvarper = 0.25,
  ncores = 1
)
tuning[c("best_nbasis", "best_r_pen", "GCV")]
```

## Run the FPCA-DE pipeline

Setting `ncores = 1` gives sequential, reproducible execution after
`set.seed()`. Parallel execution is available by increasing `ncores`.

```{r run}
set.seed(2026)
res <- scFPCDE_run(
  yt = yt,
  tt = tt,
  L = 2,
  nbasis = 30,
  r_pen = 1e-3,
  topvarper = 0.25,
  n_perm = 25,
  ncores = 1
)

names(res)
head(res$D_test_result)
```

The result contains the refined FPCA fit in `fpca_result`, the D-test table in
`D_test_result`, and `F_test_result`. The last component is `NULL` unless
`use_FPC_F = TRUE` is requested.

## Inspect differential-expression results

```{r pval}
hist(
  res$D_test_result$p_value,
  breaks = 20,
  main = "D-test permutation p-values",
  xlab = "p-value",
  col = "grey80",
  border = "white"
)
```

Because the simulation records which genes are non-null, the discoveries can be
compared with the known status. In real data, this oracle is unavailable.

```{r contingency}
is_de <- gene_index <= 100
names(is_de) <- colnames(yt)

table(
  Oracle = is_de[res$D_test_result$ID],
  Significant = res$D_test_result$q_value < 0.05
)
```

The optional FPCA F-test is enabled in the wrapper as follows:

```{r f-test, eval=FALSE}
res_with_f <- scFPCDE_run(
  yt = yt,
  tt = tt,
  n_perm = 1000,
  ncores = 2,
  use_FPC_F = TRUE
)
head(res_with_f$F_test_result)
```

## Visualize fitted trajectories

`fpca_result$xt_hat` contains fitted values for the centered expression matrix.
For a direct overlay, apply the same transformation to the observed values.

```{r plot}
yt_centered <- scale(yt, center = TRUE, scale = FALSE)
top_degs <- order(res$D_test_result$D_obs, decreasing = TRUE)[1:12]

scFPCDE_gene_curves(
  tt = tt,
  yt = yt_centered,
  yt_fit = res$fpca_result$xt_hat,
  cell_cluster = clusters,
  subset = top_degs,
  nrow = 3,
  ncol = 4
)
```

For larger analyses, first remove genes with very low detection using
`scFPCDE_filter_genes()`, tune the smoothing parameters on a representative
dataset, and save the returned result object together with the random seed and
package version used for the analysis.
